Tail associations and their impact on extinction risk

### 2019-11-03 MainText.Rmd
edited the abstract

\noindent Extreme climatic events are becoming more frequent and more intense
due to climate change. Furthermore, there is reason to believe extreme 
climatic events may modify "tail associations" between values of an environmental 
variable measured in different locations, i.e., associations that 
occur principally between values in the left or right tails of the distributions 
of values of the variable that can occur in the individual locations. 
Two positively associated variables can be "left-tail associated" (i.e., more 
correlated when they take low than high values) or "right-tail associated"
(more correlated when they take high than low values) even with the same
overall correlation coefficient in both cases. We tested whether the nature of 
spatial tail associations of environmental variables can influence metapopulation 
extinction risk. For instance, if low values of a variable are bad for local
populations, one may expect that left-tail associations increase
metapopulation extinction risks because then environmental catastrophes are
spatially synchronized, presumably reducing the potential for rescue effects.
We also tested, in the context of (non-spatial) stage-structured matrix models, 
whether tail associations between values of stage-specific vital rates may
influence extinction risk.

Tail associations and their asymmetries are taken into account in assessing risks
in finance and other fields, but to our knowledge our study is one of the first
to consider how tail associations influence extinction risk.

With global climate change, extreme climatic events (ECEs) become more frequent. These ECEs could pose a severe impact on the resistance or resilience of a population and/ or community level through life-history traits of the species. Therefore predicting the impact of extreme events on population extinction risk is necessary. We here introduce the potential use of "tail association" property of two correlated ecological variables in modeling the extinction risk of an age-structured single patch model. We also extend this idea to density-dependent metapopulation models. For metapopulation models, we show extinction risk not only depends on the nature of the tail association of spatially correlated noise but is also influenced by the inherent dynamics of the population-model. To elaborate, we demonstrate left tail dependent noise across spatially correlated patches cause higher extinction risk than the right tail dependent noise when the model show undercompensatory dynamics. But in the overcompensatory regime, the trend is the opposite. Our study paves the path for future research considering this tail association instead of a simple correlation in population viability analysis.

\noindent \textbf{\textit{Keywords:}} extinction risk, extreme events, tail association, density-dependent model

>>>>

\noindent Extreme climatic events are becoming more frequent and more intense
due to climate change. Furthermore, there is reason to believe extreme 
climatic events may modify "tail associations" between values of an environmental 
variable measured in different locations, i.e., associations that 
occur between values in the left or right tails of the distributions 
of the variable that occur in the locations. 
Two positively associated variables can be principally "left-tail associated" (i.e., more 
correlated when they take low than high values) or "right-tail associated"
(more correlated when they take high than low values), even with the same
overall correlation coefficient in both cases. We tested whether the nature of 
spatial tail associations of environmental variables can influence metapopulation 
extinction risk. For instance, if low values of a variable reduce the growth 
rates of local
populations, one may expect that left-tail associations increase
metapopulation extinction risks because then environmental catastrophes are
spatially synchronized, presumably reducing the potential for rescue effects.
We also tested, in the context of non-spatial stage-structured matrix models, 
whether tail associations between stage-specific vital rates may
influence extinction risk. 
We showed that density dependence interacts with the
propensity for tail associations to influence metapopulation extinction risk: for 
population models
showing undercompensatory density dependence, left-tail associations in environmental 
variables typically accentuated
and right-tail associations mitigated extinction risk; whereas the 
reverse was usually true for models showing overcompensatory density dependence.
For the non-spatial, stage structured models we considered, left-tail associations 
between vital rates did accentuate extinction risk compared to right-tail
associations, but the effect was small. 
Tail associations and their asymmetries are taken into account in assessing risks
in finance and other fields, but to our knowledge our study is one of the first
to consider how tail associations influence population extinction risk. 
Our modelling results 
provide an initial demonstration of a new mechanism influencing extinction risks,
and, in our view, should help motivate more comprehensive
study of the mechanism and its importance for real populations in future work.

\noindent \textbf{\textit{Keywords:}} extinction risk, extreme events, tail association, density-dependent model, copula
